Efficacy of COVID-19 vaccination against severe COVID-19 and risk to  double-vaccinated individuals in Scotland: relationship to co-morbidities conferring  clinically extremely vulnerable status or moderate risk

### 2021-08-13 vaxshieldupdate.Rmd
now loads cc.all

source("vaxshieldupdate.R")

@@

The effect of vaccination in each of the clinical vulnerability categories was estimated in a conditional logistic regression model (R function `survival::clogit`, package version_3.2-7, R version 3.6.3) fitted to the full dataset, specifying effects \ensuremath{\beta_{R2}, \dots, \beta_{RJ}} for the log rate ratio associated with risk categories 2 to \ensuremath{J} (\ensuremath{\beta_{R1} = 0} for the reference category \ensuremath{J=1}), and nested effects \ensuremath{\beta_{V1}, \dots, \beta_{VJ}} for the log rate ratio associated with vaccination in each of the \ensuremath{J} risk categories. With this incidence density sampling design, the conditional odds ratio is the rate ratio.  The efficacy of vaccination is 1 minus the rate ratio.  The strata in the conditional logistic regression model are matched sets: typically one case and up to 10 controls, but where two cases with the same age, sex and GP practice present on the same day the matched set comprises two cases and up to 20 controls.  When estimating the rate ratio within a risk group such as a shielding category, only cases and controls in that risk group are retained in the matched sets.   Matched sets in which all are vaccinated or none are vaccinated do not contribute to the rate ratio.   The unconditional odds ratios calculated from frequency tables of the vaccination status of cases and controls in each risk group cannot be used to estimate rate ratios [@breslow_estimation_1978;@greenland_need_1982].   The design controls not only for the matching factors of age, sex and general practice but also for calendar time.  Other variables included as covariates in the models had been previously identified as strongly associated with severe COVID-19: care home residence, number of adults in the household, number of distinct drug classes (BNF subparagraph codes) for which prescriptions were dispensed between 15 and 240 days before presentation date, and any hospital in-patient stay from 5 to 14 days before presentation date.

# Cohort analysis of the shielding list
The case-control study can estimate only rate ratios.  To investigate the absolute rates of severe disease in those listed as clinically extremely vulnerable, we undertook a separate cohort study of all individuals who had ever been on the shielding list. At the time of this study  the dates of addition and removal from the shielding list, and linkage of the full cohort to vaccination status and mortality were not available to us, so this analysis is only provisional.  A Poisson regression model was fitted to a dataset formatted with one observation per 28-day person-time interval, and individuals censored at first event.  The event was defined as severe COVID-19 presenting within the person-time interval, and the covariates were sex, age, and shielding eligibility group.  The baseline hazard rate was specified as a natural spline function with 4 degrees of freedom.  Similar models were fitted with event indicator defined as hospitalized or fatal cases, or as any diagnosed case.

# Results
## Time course of the epidemic and associations with risk factors 
Fig \ref{fig:plotcases} shows the daily numbers of severe cases (averaged over 3-day sliding windows) by risk category.  From mid-January to early March 2021 there was a rapid decline in the incidence of severe cases, followed by a slight increase from early June to early July.

Fig \ref{fig:plotrateratios} shows the time course of rate ratios (computed over 42-day sliding windows) for risk category, vaccination dose (coded as 0, 1, 2 doses), and recent (5 to 14 days before presentation date) hospital exposure as inpatient.  These rate ratios are computed from a model that includes all variables shown, together with care home residence.

The rate ratio for severe disease associated with shielding eligibility did not decline as the winter peak declined, and increased from May 2020 onwards.  The efficacy of vaccination against severe disease (modelled as the rate ratio per 2 doses) remained high, despite the rapid replacement of the Alpha by the Delta variant during May 2020.

@@

# Discussion
## Statement of principal findings

1. Although the efficacy of vaccination against severe disease in those without risk conditions remains high (at least 90%), it is now clear that the efficacy in clinically extremely vulnerable individuals is considerably lower (about 70%).  There is however a dose-response effect in this group: two doses are more effective than one.

@@

knitr::kable(table.shieldgr[14:21, ], 
             escape=FALSE, 
             booktabs=TRUE,
			 label="rateratiosshieldgroups",
             row.names=FALSE,
             col.names=c("Effect", "Rate ratio (95\\% CI)", "\\ensuremath{p}-value"),
			 align=c('l', 'c', 'c'), 
             caption="Rate ratios for severe COVID-19 within shielding eligibility groups: vaccine dose encoded as numeric variable taking values 0, 1/2, 1") %>%
add_footnote(label=c("Presentation dates from 1 December 2020 to 14 July 2021.",
                     "Severe COVID-19 defined as entry to critical care or death within 28 days of presentation.", 
                     "Controls matched for age, sex, and primary care practice on date of presentation of case.",
                     "Adjusted for care home residence, number of adults in household, number of drug classes and recent hospital stay as covariates", 
                     "Rate ratios estimated in multivariable conditional logistic regression with all variables in model",  
                     "Vaccination status is number of doses administered at least 14 days before presentation date"), 
             notation="none") %>%
    column_spec(1, width="8cm") %>%
column_spec(2, width="4cm") %>%
column_spec(3, width="3cm") %>%
    kable_styling(latex_options=c("HOLD_position"))

@@

```{r plotrateratios, echo=FALSE, warning=FALSE, message=FALSE, fig.pos = "H", fig.width=6, fig.asp=0.8, fig.cap="\\label{fig:plotrateratios} Rate ratios by 42-day sliding window. For each time window, the conditional logistic regression model includes all variables shown in the plot, together with care home residence. Line thickness is inversely proportional to standard error"}

@@

knitr::kable(table.hosp.shieldgr[13:20, ], 
             escape=FALSE, 
             booktabs=TRUE,
			 label="hosprateratiosshieldgroups",
             row.names=FALSE,
             col.names=c("Effect", "Rate ratio (95\\% CI)", "\\ensuremath{p}-value"),
			 align=c('l', 'c', 'c'), 
             caption="Rate ratios for hospitalized or fatal COVID-19 within shielding eligibility groups: vaccine dose encoded as numeric variable taking values 0, 1, 2") %>%
add_footnote(label=c("Presentation dates from 1 December 2020 to 14 July 2021.",
                     "Controls matched for age, sex, and primary care practice on date of presentation of case.",
                     "Rate ratios estimated in multivariable conditional logistic regression with all variables in model",  
                    "Vaccination status is number of doses administered at least 14 days before presentation date"), 
             notation="none") %>%
    column_spec(1, width="8cm") %>%
column_spec(2, width="4cm") %>%
column_spec(3, width="3cm") %>%
    kable_styling(latex_options=c("HOLD_position"))

>>>>

@@

The effect of vaccination in each of the clinical vulnerability categories was estimated in a conditional logistic regression model (R function `survival::clogit`, package version_3.2-7, R version 3.6.3) fitted to the full dataset, specifying effects \ensuremath{\beta_{R2}, \dots, \beta_{RJ}} for the log rate ratio associated with risk categories 2 to \ensuremath{J} (\ensuremath{\beta_{R1} = 0} for the reference category \ensuremath{J=1}), and nested effects \ensuremath{\beta_{V1}, \dots, \beta_{VJ}} for the log rate ratio associated with vaccination in each of the \ensuremath{J} risk categories. With this incidence density sampling design, the conditional odds ratio is the rate ratio.  The efficacy of vaccination is 1 minus the rate ratio.  The strata in the conditional logistic regression model are matched sets: typically one case and up to 10 controls, but where two cases with the same age, sex and GP practice present on the same day the matched set comprises two cases and up to 20 controls.  When estimating the rate ratio within a risk group such as a shielding category, only cases and controls in that risk group are retained in the matched sets.   Matched sets in which all are vaccinated or none are vaccinated do not contribute to the rate ratio.   The unconditional odds ratios calculated from frequency tables of the vaccination status of cases and controls in each risk group cannot be used to estimate rate ratios [@breslow_estimation_1978;@greenland_need_1982].   The design controls not only for the matching factors of age, sex and general practice but also for calendar time.  Other variables included as covariates in the models had been previously identified as strongly associated with severe COVID-19: care home residence, number of adults in the household, number of distinct drug classes (BNF subparagraph codes) for which prescriptions were dispensed between 15 and 240 days before presentation date, and any hospital in-patient stay from 5 to 14 days before presentation date. 
<!---
# Cohort analysis of the shielding list
The case-control study can estimate only rate ratios.  To investigate the absolute rates of severe disease in those listed as clinically extremely vulnerable, we undertook a separate cohort study of all individuals who had ever been on the shielding list. At the time of this study  the dates of addition and removal from the shielding list, and linkage of the full cohort to vaccination status and mortality were not available to us, so this analysis is only provisional.  A Poisson regression model was fitted to a dataset formatted with one observation per 28-day person-time interval, and individuals censored at first event.  The event was defined as severe COVID-19 presenting within the person-time interval, and the covariates were sex, age, and shielding eligibility group.  The baseline hazard rate was specified as a natural spline function with 4 degrees of freedom.  Similar models were fitted with event indicator defined as hospitalized or fatal cases, or as any diagnosed case.  
--->

# Results
## Time course of vaccine efficacy against severe disease 
Fig \ref{fig:plotcases} shows the daily numbers of severe cases (averaged over 3-day sliding windows) by risk category.  In the first half of July the average daily number of severe cases among the 200,000 people still on the shielded list was about 4, equating to an incidence rate of about 0.6 per 1000 per month.

Fig \ref{fig:plotrateratios} shows the time course of rate ratios (computed over 42-day sliding windows) for risk category and vaccination status, in a model that includes both these variables together with care home residence and recent hospital stay as covariates.  This shows that the rate ratio for severe disease associated with shielding eligibility, adjusted for vaccination status, has increased since 1 May, even though the absolute rate of severe disease in this group remains much lower than in December-January.  The rate ratios for severe disease associated with 1 dose and 2 doses of vaccine at least 14 days earlier (with unvaccinated as reference category) declined  steadily from January to June, implying that the efficacy increased despite the rapid replacement of the Alpha variant by the Delta variant during May 2020 [@sheikh_sarscov2_2021].

@@

# Discussion
## Statement of principal findings
1. Although the efficacy of vaccination against severe disease in those without risk conditions remains high (at least 90%), it is now clear that the efficacy in clinically extremely vulnerable individuals is considerably lower (about 70%).  There is however a dose-response effect in this group: two doses are more effective than one.

@@

table.shieldgr[, effect := gsub("vax", "vaccine", effect)] 
knitr::kable(table.shieldgr[14:19, ], 
             escape=FALSE, 
             booktabs=TRUE,
			 label="rateratiosshieldgroups",
             row.names=FALSE,
             col.names=c("Effect", "Rate ratio (95\\% CI)", "\\ensuremath{p}-value"),
			 align=c('l', 'c', 'c'), 
             caption="Rate ratios for severe COVID-19 within shielding eligibility groups: vaccine dose encoded as numeric variable taking values 0, 1/2, 1") %>%
add_footnote(label=c("Presentation dates from 1 December 2020 to 14 July 2021.",
                     "Severe COVID-19 defined as entry to critical care or death within 28 days of presentation.", 
                     "Controls matched for age, sex, and primary care practice on date of presentation of case.",
                     "Adjusted for care home residence, number of adults in household, number of drug classes and recent hospital stay as covariates", 
                     "Rate ratios estimated in multivariable conditional logistic regression with all variables in model",  
                     "Vaccination status is number of doses administered at least 14 days before presentation date"), 
             notation="none") %>%
    column_spec(1, width="8cm") %>%
column_spec(2, width="4cm") %>%
column_spec(3, width="3cm") %>%
    kable_styling(latex_options=c("HOLD_position"))

@@

```{r plotrateratios, echo=FALSE, warning=FALSE, message=FALSE, fig.pos = "H", fig.width=6, fig.asp=0.8, fig.cap="\\label{fig:plotrateratios} Rate ratios by 42-day sliding window. For each time window, the conditional logistic regression model includes all variables shown in the plot, together with care home residence and recent hospital stay. Line thickness is inversely proportional to standard error"}

@@

table.hosp.shieldgr[, effect := gsub("vax", "vaccine", effect)] 
knitr::kable(table.hosp.shieldgr[14:19, ], 
             escape=FALSE, 
             booktabs=TRUE,
			 label="hosprateratiosshieldgroups",
             row.names=FALSE,
             col.names=c("Effect", "Rate ratio (95\\% CI)", "\\ensuremath{p}-value"),
			 align=c('l', 'c', 'c'), 
             caption="Rate ratios for hospitalized or fatal COVID-19 within shielding eligibility groups: vaccine dose encoded as numeric variable taking values 0, 1, 2") %>%
add_footnote(label=c("Presentation dates from 1 December 2020 to 14 July 2021.",
                     "Controls matched for age, sex, and primary care practice on date of presentation of case.",
                     "Rate ratios estimated in multivariable conditional logistic regression with all variables in model",  
                    "Vaccination status is number of doses administered at least 14 days before presentation date"), 
             notation="none") %>%
    column_spec(1, width="8cm") %>%
column_spec(2, width="4cm") %>%
column_spec(3, width="3cm") %>%
    kable_styling(latex_options=c("HOLD_position"))
